Title Goes Here

### 2023-01-04 inst/paper/_discussion.qmd
add: revisions based on co-authors feedback

<!-- Big summary/important statement for the science -->
In this study we were able to extract and identify multiple alkaloids from the
dental calculus of individuals from Middenbeemster, a 19th century Dutch
archaeological site. We applied ultra-high-performance liquid chromatography-tandem
mass spectrometry (UHPLC-MS/MS), using a method that was validated the presence
of drugs and metabolites in dental calculus with the equivalent compounds in blood
[@sorensenDrugsCalculus2021].
Here we have shown that the method can also be successfully applied to archaeological
dental calculus.
We initially detected cannabinoids, cocaine, nicotine, cotinine, caffeine,
theopyhlline, and salicylic acid in multiple individuals; however, the detection
of cannabinoids and cocaine could not be replicated in a second run on different
samples from the same individuals.

<!-- discuss any correlations, spurious or otherwise
e.g. more caffeine, salicyl -> less calculus (acidic environ?)
tobacco -> sinusitis -->

<!-- compounds present and absent with potential sources-->

<!-- nicotine and cotinine: tobacco, obviously...-->
Nicotine and cotinine were strongly positively correlated, both in concentration
and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
The detection of nicotine and cotinine is not surprising, as pipe-smoking in
the Beemsterpolder is well-documented in the literature
[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
remains [@lemmersMiddenbeemster2013].
There is also documented medicinal use of nicotine in the Beemsterpolder,
where a tobacco
smoke enema was used for headaches, respiratory problems, colds, and drowziness,
among others, from around 1780 to 1830 [@aten400JaarBeemster2012].
<!-- Scientific findings as extension of previous work -->
The detection of nicotine in dental calculus has previously been presented by
Eerkens and colleagues [-@eerkensDentalCalculus2018] in two individuals from
pre-contact California. They also targeted caffeine, cotinine, and theophylline
in their samples, but were unable to detect any of them. It remains to be seen
whether this is due to differences in methods used, or due to our samples being
more recent.
Other alkaloids have been detected in archaeological samples, though these are more
often associated with other materials, including pipes
[@raffertyCurrentResearch2012] and ceramics [@smithDetectionOpium2018].
Direct analysis on human remains
has also included hair [@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013].

<!-- theophylline and caffeine: coffee and tea? -->
Theophylline and caffeine concentrations were positively correlated,
suggesting that they may have originated from the same source.
Caffeine is present in coffee, tea, and cocoa beans, with concentrations slightly higher
in coffee [@stavricVariabilityCaffeine1988; @chinCaffeineContent2008; @bispoSimultaneousDetermination2002].
Theophylline is present in both coffee beans and tea leaves, but in negligible
quantities [@stavricVariabilityCaffeine1988]. It is also metabolised from
caffeine in the liver, and given the low quantity relative to caffeine, it likely
originated from either tea or coffee, which contain higher concentrations of
the compounds [@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
Caffeine and theophylline are
structurally very similar, and we expect they would experience similar rates of
degradation, allowing us to make interpretations on the ratio between the compounds.
Tea consumption had become widespread in the Netherlands by 1820, reaching all aspects of
the population (some argue that this happened already by early 18th century)
[@nierstraszTeaTrade2015, p.91].
Historically, we also know that both tea and coffee were consumed in the Beemsterpolder
during the 19th century. 'Theegasten' (teatime) was a special
occasion occurring from 15.00-20.00 hours, where tea was served along with the evening
bread [@schuijtemakerTeTheegasten2011]. Many households also owned at least one coffee pot and tea pot [@boumanBegravenis2017].
Both the presence of theophylline and caffeine could also be explained by chocolate,
although given the large quantities of caffeine compared to theophylline, it is more
likely that the source is tea and/or coffee
[@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
Distinguishing between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is present in
higher quantities in chocolate compared to caffeine and theophylline
[@stavricVariabilityCaffeine1988; @bispoSimultaneousDetermination2002; @alanonAssessmentFlavanol2016].
Although there is some indication that theobromine does not preserve well
[@velskoDentalCalculus2017], and frequent consumption of all three
items would be difficult to parse.

<!-- salicylic acid: diet, medicine, or environmental??? -->
The ubiquitous presence of salicylic acid poses a different problem. It has
been shown that salicilyc acid is a very mobile organic acid and the ubuquitous
presence may be due to environmental contamination
[@badriRegulationFunction2009; @velskoDentalCalculus2017].
The bark of a willow tree, *Salix alba*, has long been known for its
pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119],
and is also present in many plant-based foods
[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017]. This includes
potatoes, which were a mainstay in the Beemsterpolder diet [@aten400JaarBeemster2012].
It will be necessary to explore the extent to which salicylic acid can leech
into the dental calculus from the soil, and what the rate of degradation is for
salicylic acid when trapped in dental calculus.

<!-- cannabinoids: medicinal? wait, where did they go..? -->
Cannabinoids---specifically THC, THCA-A, THCVA, CBD, CBN---were found in the
first batch, but none were replicated in the second batch. Medicinal use of
cannabinoids has been well-established in Europe since Medieval-times, and it
was also grown in the Netherlands [@bruinsmaBijdragenTot1872].
Administration was most common in the form of concoctions containing various portions
of the cannabis plant for ingestion; not until the late 19th century did it become
recommended to smoke it for more immediate effects [@clarkeCannabisEvolution2013].
John Parkinson described a Dutch medicinal use of hemp. An emulsion was prepared
from the seeds of the plants to treat pain and various stomach ailments. Another
preparation involving the roots of the plants was used for inflammation, gout, and joint
pains [@clarkeCannabisEvolution2013].
The ability to detect cannabinoids in calculus may be limited by a reduced ability
to diffuse from serum to salivary glands due to an affinity for protein-binding,
[@sorensenEffectAntioxidants2018; @sorensenDrugsCalculus2021],
meaning detection would rely on oral consumption of the cannabinoids. Even then,
the overall instability of some cannabinoids could also affect detection
[@sorensenEffectAntioxidants2018; @lindholstLongTerm2010].

<!-- opioids: medicinal. Damn... that would've been cool-->
Despite many of the individuals living during the height of the opium era in the
Netherlands, none of the targeted opioids
(morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
were detected. 
Sylvius de la Boe (1614–1672), a German-born physician who studied anatomy at
Leiden University, the Netherlands, relied heavily on opium [@machtHistoryOpium1915].
The absence of opioids could be a result of the people ascribing more to the
"traditional" rather than "scientific" medicine, although laudanum and
another opium containing concoction was part of the "traditional" medicine in
the Netherlands [@leuwProhibitionLegalizationDutch1994], including Middenbeemster
[@aten400JaarBeemster2012]. It was also generally
considered a drug of the upper class [@scheltemaOpiumTrade1907], and may be
more common in an urban center.
The absence could also be attributed to postmortem degradation.
It has been shown that, while abundant in opium, morphine
degrades rapidly. The more stable opioids such as thebaine and papaverine
[@chovanecOpiumMasses2012] were also absent, however.
<!--Similar to cannabinoids, many opioids are preferentially bound to (plasma) proteins
[@sorensenDrugsCalculus2021].-->

We explored the relationship between detected compounds and various skeletal
indicators, such as pathological and dental lesions, preservation, and pipe notches.
We found no evidence to suggest that preservation of the skeleton influences
the recovery of compounds from the dental calculus. This could be due to
the good overall preservation of both teeth and dental calculus, which often
preserves better than the rest of the skeleton.
We found no correlation between the number of pipe notches
and the concentration of nicotine and cotinine, suggesting that our ability
to detect tobacco consumption in dental calculus does not necessarily rely
targeting frequent smokers; here, we consider individuals with multiple pipe notches
as likely to have been heavy smokers.
There does seem to be a positive correlation between the weight of the calculus
sample and recovered quantities of the compounds, so weight of the sample may affect
the ability to detect compounds; although, we were able to detect
compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].

[[ Correlation between caffeine, nicotine, and sinusitis - Maia ]]

Authentication

<!-- authentication -->
The list of targeted compounds included both naturally occurring compounds known
to have been used in the past, as well as synthetic modern drugs did not exist
at the time (e.g. Fentanyl, MDMA, Amphetamine). These were part of the toxicology
screening for the original method [@sorensenDrugsCalculus2021], originally
developed on cadavers. In our study they serve as an authentication step, as their
presence in archaeological samples could only be the result of contamination.
<!-- the importance of replication - brief discussion of cocaine -->
The only modern compound (at least in a European context) detected in the sample
was cocaine, which was detected in the first batch of sampels. The population
analysed is from the early-mid 18th century, and cocaine was isolated in 1860 by
Albert Niemann, and entered popular medical
practice in 1884. Coca arrived in Europe as early as 1771, but as botanical
specimens rather than for consumption, and there were also issues importing enough
viable specimens of coca for cocaine extraction
[@mortimerHistoryCoca1901 p. 179; @abucaCocaTrade2019, p.108].
We considered it possible that it would be present in a sample with
most individuals originating from the early- to mid-19th century. If corroborated,
this would have been the first case of coca-consumption in Europe. Given that cocaine
had yet to be isolated, the only way it could have gotten into the calculus was
through consumption of the coca leaf. In our replication batch, we included all
of the individuals who had been cocaine-positive in the first batch. We were
unable to replicate any of the cocaine results, and we were unable to detect the
principal metabolite, benzoylecgonine, in either batch; although, the original
study only detected trace amounts of benzoylecgonine, and only in some of the
individuals that were cocaine-positive. We hope this will serve as
a reminder of the importance of replication. We suspect that the original detection
of cocaine was a result lab contamination during the UHPLC-MS/MS analysis.

Accuracy

We used the presence of a pipe notch and concurrent detection of nicotine
and/or cotinine was as a crude estimate of the accuracy of the method. Nicotine
has been widely studied in the archaeological record due to its apparent stability
and ability to survive over long periods of time
[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].
The accuracy was found to be around
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
We acknowledge that this is
a very rough estimate, as the presence of a pipe notch is likely not a perfect
indicator of whether or not someone consumed tobacco, and it may not be
representative of other compounds with different rates of degradation.

@@

<!-- secondary findings-->
The choice of tooth for sampling may impact the detection of certain compounds,
as the incorporation in dental calculus may depend on the mode of consumption.
Tobacco smokers may have more
nicotine present in calculus on incisors, whereas tobacco chewers may have more
on molars [@eerkensDentalCalculus2018]. This has yet to be systematically
tested in archaeological dental calculus. Direct incorporation by ingesting
items, food or otherwise, containing the compounds
is not the only potential pathway to incorporation in the dental calculus. 
The presence of cotinine, the primary
metabolite of nicotine, suggests that saliva is also a source of deposition, and
that deposition of alkaloids in dental calculus can occur both on the way into the body,
i.e. during consumption, and on the way out, i.e. disposal of waste products via
saliva secretion into the mouth. Especially mucin-rich saliva from the sublingual
and submandibular glands preferentially binds toxins [@doddsHealthBenefits2005],
and since these glands are located closest to the lower incisors, they may be the
most effective target for these studies. Due to homogenisation of samples in our
study, we were unable to test this effect.

<!-- How this engineering will facilitate future research -->
<!-- both calculus and teeth preserve well-->

@@

<!-- limitations of the platform show which questions would be unanswered -->
Limitations

Since we targeted individuals with moderate-to-large calculus deposits,
it is likely a biased sample. The presence of calculus may increase risk of premature death
[@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
be associated with dental calculus build-up) is a risk-factor for
respiratory diseases, if periodontal and respiratory pathogens
enter the blood stream
[@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
In our sample, the percentage of CMS
(`r scales::percent(mean(explore_select$CMS, na.rm = T), accuracy = 0.1)`)
is lower than in another (more representative) male sample (44.1%)
[@casnaUrbanizationRespiratory2021], and the caries percentage is similarly lower
in our sample (`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`) than
a more representative sample (22.9%) [@lemmersMiddenbeemster2013].
The bias may also affect the detection of compounds, as caffeinated drinks are often
acidic, and may cause a lower calculus formation<!-- citation?-->.
We only found a weak negative correlation (`r signif(corr_tib[15,2], 3)`)
between caffeine concentration and dental calculus index, but since we targeted individuals with
calculus, this may inadvertently have been controlled for during sample collection.

Quantitation of the detected compounds may have limited value in archaeological
samples, given the effects of taphonomy added on top of individual variability
and relative stochasticity of dental calculus formation, and the metabolic
degradation of compounds during life [@sorensenDrugsCalculus2021; @velskoAncientDental2022a]. <!-- the results of comparisons with concentration
had limited outcomes, and the correlations between concentration and other variables
were similar to the correlations with presence/absence, suggesting that the actual
concentration was somewhat random.-->

Perhaps the main limitation, which is also true for many other analyses on
archaeological materials, is that detecting the absence of a compound is not possible.
The absence of a compound is
not evidence of its absence, which complicates the interpretation of our results.
We have attempted to minimise errors occurring due to this limitation by including
a relatively large sample of individuals. Although given the relatively low
detection rate seen in tobacco, this remains a major limitation, and will likely
been compounded by increasing antiquity of the samples.
The replication of samples increases the evidence of absence if the compound
was absent in both batches.

<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->

While a major limitation is the uncertainty surrounding whether or not a compound
is actually absent, the power of the method is in detecting food and other items
that are consumed in a manner that does not allow macroscopic detection. Taking
tobacco consumption as an example; while pipe
notches are a useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, which includes chewing, drinking, cigars, snuff,
[@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
was mostly common in males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018].
Our method does not, at least to our knowledge, discriminate between modes of
consumption,
and may be more suitable for exploring tobacco consumption in the entire population,
not just males. In our sample, we detected nicotine and cotinine (replicated)
in an old adult, probable female individual. Because the individual was an old
adult, it is unlikely that the compounds entered the dental calculus through
pipe-smoking, as there would be pipe notches. More likely that the tobacco entered
through alternative consumption or secondhand smoke.

Future studies should explore the effect of tooth position on the detection of
compounds.
Also on the properties increasing incorporation potential of different compounds,
and which pathways facilitate maximum incorporation (direct or indirect pathways?),
which likely differs between alkaloids. How is the preservation once trapped
inside the calculus, for example using artificial ageing on real or artificial
dental calculus?
In this study, we only targeted drugs that were included in the forensic toxicological
screenings, and therefore only covered a limited number of the potential compounds
that could be of interest for exploring past diets and medicinal treatments.

<!--final remarks-->

This preliminary study outlines the benefits of using calculus to target a
variety of compounds that could have been consumed as medicine or diet. 
This method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in historic
documentation, such as lower status, children and women.
We also
show that there are many limitations that will need to be addressed going forward
with this type of analysis, and stress the need for more systematic research on
the consumption of alkaloid-containing items and their subsequent concentration
and preservation in dental calculus,
in addition to how mode of consumption may affect concentrations on different parts
of the dentition. <!-- also individual variation -->
Another limitation of dental calculus as a medium is the variability its formation
and the many factors that can influence incorporation and retention of molecules
and particles; however, in the absence
of hair and serum (quite uncommon in archaeology), dental calculus represents
an impressive reservoir of information regarding the consumption of various
alkaloids, whether dietary, medicinal, recreational, or otherwise.

>>>>

<!-- Big summary/important statement for the science -->
In this study we were able to extract and identify multiple alkaloids from the
dental calculus of individuals from Middenbeemster, a 19th century Dutch
archaeological site. We applied ultra-high-performance liquid chromatography-tandem
mass spectrometry (UHPLC-MS/MS), using a method that was validated by co-occurrence
of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
Here we have shown that the method can also be successfully applied to archaeological
dental calculus.
We extend findings from previous studies on alkaloids in archaeological samples
by extracting multiple different alkaloids from dental calculus, including
cannabinoids, cocaine, nicotine, cotinine, caffeine, theophylline, and salicylic
acid in multiple individuals. Of these compounds, the cannabinoids and cocaine
could not be replicated in a second run on different samples from the same individuals.
We discuss the implications of these findings in light of historical and
archaeological evidence for the consumption of these drugs.

<!-- compounds detected with potential sources, and those that were surprisingly missing -->

<!-- nicotine and cotinine: tobacco, obviously...-->
Nicotine and its principal metabolite, cotinine, were strongly positively correlated,
both in concentration and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
The detection of nicotine and cotinine is not surprising, as tobacco-use and
pipe-smoking in the Beemsterpolder is well-documented in the literature
[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
remains as pipe notches [@lemmersMiddenbeemster2013].
There is also documented medicinal use of nicotine in the Beemsterpolder,
where a tobacco-smoke enema was used for headaches, respiratory problems, colds,
and drowsiness from around 1780 to 1830 [@aten400JaarBeemster2012].
In our sample, we also detected nicotine and cotinine (replicated)
in an old adult, probable female individual.
It is unlikely that the compounds entered the dental calculus through
pipe-smoking, as the individual had no visible pipe notches.
More likely that the tobacco entered through an alternate mode of consumption,
secondhand smoke, or the aforementioned tobacco-smoke enema.

<!-- Scientific findings as extension of previous work -->

<!-- strong correlation between nicotine and cotinine concentration is important?
      Could suggest that the incorporation pathway is excretion, not consumption
        - longer time in body than mouth (consumption)
-->

<!-- theophylline and caffeine: coffee and tea? -->
Theophylline and caffeine were positively correlated in our samples, though to a
lesser extent than nicotine and cotinine, so we are unable to determine if they
originated from the same source (@tbl-pearson and @fig-polycorr).
Caffeine and theophylline have very similar chemical structures, so we expect
they would experience similar rates of incorporation and degradation
degradation, allowing us to make interpretations the ratio and correlations
between the compounds.
Caffeine is present in coffee, tea, and cocoa beans, with concentrations slightly
higher in coffee
[@stavricVariabilityCaffeine1988; @chinCaffeineContent2008; @bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
Theophylline is present in both coffee beans and tea leaves, but in negligible
quantities [@stavricVariabilityCaffeine1988]. It is also a primary metabolite of
caffeine produced by the liver. Given the low correlation, there are likely
multiple sources of caffeine and theophylline in the population, with tea and
coffee being the most obvious.  
Tea consumption had become widespread in the Netherlands by 1820, reaching all
parts of society [@nierstraszTeaTrade2015, p.91].
Historically, we also know that both tea and coffee were consumed in the
Beemsterpolder during the 19th century. 'Theegasten' (teatime) was a special
occasion occurring from 15.00-20.00 hours, where tea was served along with the evening
bread [@schuijtemakerTeTheegasten2011]. Many households also owned at least one
coffee pot and tea pot [@boumanBegravenis2017].
Distinguishing between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is present in
higher quantities in chocolate compared to caffeine and theophylline
[@stavricVariabilityCaffeine1988; @bispoSimultaneousDetermination2002; @alanonAssessmentFlavanol2016].
However, In addition to oral factors
affecting alkaloid uptake in dental calculus, there is some indication that
theobromine does not preserve well in the archaeological record
[@velskoDentalCalculus2017], and frequent consumption of all three items would
be difficult to parse.

<!-- salicylic acid: diet, medicine, or environmental??? -->
Salicylic acid was found in all but one individual in our sample.
It can be extracted from the bark of willow trees, *Salix alba*, and
has long been used for its
pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119]. It is also present
in many plant-based foods
[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017], including
potatoes, which were a staple of the Beemsterpolder diet [@aten400JaarBeemster2012].
The extracted quantity from our samples decreased over the three washes, followed
by a sharp increase in the final calculus extraction, which is what we would expect
to see if the salicylic acid was incorporated during life [@fig-auth-plot-batch2].
However, it has been shown that salicilyc acid is a very mobile organic acid and
the ubiquitous presence may be due to environmental contamination, which would
also explain the high quantity in the washes
[@chenCa2Dependent2001; @badriRegulationFunction2009].
Given the multiple plausible sources of this residue, it will be necessary to
explore the extent to which salicylic acid can leach
into the dental calculus from the soil, and what the rate of degradation is for
salicylic acid when trapped in dental calculus.
<!-- perhaps using an oral biofilm model ;) -->

<!-- cannabinoids: oh cool! medicinal? wait, where did they go..? -->
Cannabinoids---specifically THC, THCA-A, THCVA, CBD, CBN---were found in the
first batch, but none were replicated in the second batch. Medicinal use of
cannabinoids has been well-established in Europe since Medieval-times, and it
was also grown in the Netherlands [@bruinsmaBijdragenTot1872].
Administration was most common in the form of concoctions containing various portions
of the cannabis plant for ingestion; not until the late 19th century did it become
recommended to smoke it for more immediate effects [@clarkeCannabisEvolution2013].
A Dutch medicinal use of hemp involved an emulsion prepared from the seeds of
the plants to treat pain and various stomach ailments. Another
preparation involving the roots of the plants was used for inflammation, gout,
and joint pains [@clarkeCannabisEvolution2013].
The ability to detect cannabinoids in calculus may be limited by a reduced ability
to diffuse from serum to salivary glands due to an affinity for protein-binding,
[@sorensenEffectAntioxidants2018; @sorensenDrugsCalculus2021],
meaning detection would rely on oral consumption of the cannabinoids. Even then,
the overall instability of some cannabinoids could also affect detection
[@sorensenEffectAntioxidants2018; @lindholstLongTerm2010].
However, given the lack of replication, we cannot with security confirm that
cannabis was used by the Beemster population.

<!-- opioids: Damn... that would've been awesome -->
Despite many of our sampled individuals living during the height of the opium era
in the Netherlands, none of the targeted opioids
(morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
were detected. 
Sylvius de la Boe (1614–1672), a German-born physician who studied anatomy at
Leiden University, the Netherlands, relied heavily on opium [@machtHistoryOpium1915].
The absence of opioids could be a result of the people ascribing more to the
"traditional" rather than "scientific" medicine, although laudanum and
another opium containing concoction was part of the "traditional" medicine in
the Netherlands [@leuwProhibitionLegalizationDutch1994], including Middenbeemster
[@aten400JaarBeemster2012]. It was also generally
considered a drug of the upper class [@scheltemaOpiumTrade1907], and may be
more common in an urban center.
The absence could also be attributed to postmortem degradation.
It has been shown that, while abundant in opium, morphine
degrades rapidly, while thebaine and papaverine are more resistant to various
ageing processes [@chovanecOpiumMasses2012]. The latter were also absent from
our samples.

<!-- authentication -->
<!-- the importance of replication - brief discussion of cocaine -->
The only strictly modern compound (at least in a European context) detected in the sample
was cocaine, which was detected in the first batch of samples. Our sample is
derived from an early--mid 19th century population, and cocaine was isolated in 1860
by Albert Niemann, and entered popular medical
practice in 1884. Coca arrived in Europe as early as 1771, but as botanical
specimens rather than for consumption, and there were also issues importing enough
viable specimens of coca for cocaine extraction
[@mortimerHistoryCoca1901 p. 179; @abucaCocaTrade2019, p.108].
We considered it possible that it would be present in a sample with
most individuals originating from the early- to mid-19th century. If corroborated,
this would have been the first case of coca-leaf-consumption in Europe.
In our replication batch, we included all
of the individuals who had been cocaine-positive in the first batch. We were
unable to replicate any of the cocaine results, and we were unable to detect the
principal metabolite, benzoylecgonine, in either batch.
We suspect that the original detection of cocaine was a result of lab contamination
during analysis.<!--and hope this will serve as a reminder of the importance of replication.-->

<!-- correlations with skeletal indicators -->
We explored the relationship between detected compounds and various skeletal
indicators, such as pathological and dental lesions, preservation, and pipe notches.
We found no evidence to suggest that preservation of the skeleton influences
the recovery of compounds from the dental calculus. This could be due to
the good overall preservation of both teeth and dental calculus, which often
preserves better than the rest of the skeleton.  
We found no correlation between the number of pipe notches
and the concentration of nicotine and cotinine, suggesting that our ability
to detect tobacco consumption in dental calculus does not necessarily rely on
targeting frequent smokers; here, we consider individuals with multiple pipe notches
as likely to have been heavy smokers.  
<!-- inter-individual variability in calculus formation and degradation depending
on microbiome, diet, etc. -->
We found a positive correlation between CMS and nicotine, which may be
indicative of the impact tobacco smoking had on the respiratory health of the
Beemster inhabitants.
Tobacco smoke may play a significant role in diseases
of the upper respiratory tract, including CMS [@rehImpactTobacco2012]. Although
the mechanisms by which smoking increases the risk of infections is not fully
understood, solid evidence has been presented linking tobacco smoke to
increased mucosal permeability and impairment of mucociliary clearance
[@arcaviCigaretteSmoking2004]. Such changes, together with an altered immunologic
response, are thought to predispose to the development of CMS
[@slavinDiagnosisManagement2005].  
We also observed a moderate positive correlation between CMS and caffeine which
contradicts previous research linking chronic coffee consumption with a positive
effect on the respiratory system, suggesting a preventive association between
caffeine intake and pneumonia
[e.g. @alfaroChronicCoffee2018; @kondoAssociationCoffee2021].
However, while the
upper respiratory tract seems to benefit from chronic coffee consumption, it is
possible that elevated caffeine intake impacts mucosal moisture due to its
dehydrating effect [@maughanCaffeineIngestion2003], thereby exposing individuals
to greater risk of sinus infection.

@@

<!-- secondary findings-->

The detection of nicotine in dental calculus has previously been presented by
Eerkens and colleagues [-@eerkensDentalCalculus2018] in two individuals from
pre-contact California. They also targeted caffeine, cotinine, and theophylline
in their samples, but were unable to detect any of them. It remains to be seen
whether this is due to differences in methods used, or due to our samples being
more recent.
They also suggest that the choice of tooth for sampling may impact the detection
of certain compounds,
as the incorporation in dental calculus may depend on the mode of consumption.
Tobacco smokers may have more
nicotine present in calculus on incisors, whereas tobacco chewers may have more
on molars [@eerkensDentalCalculus2018].
However, sampling may not be limited to mode of consumption.
The presence of cotinine suggests that the excretion of a compound after being
metabolised in the body is also a source of deposition, and
that deposition of alkaloids in dental calculus can occur both on the way into
the body, i.e. during consumption, and on the way out, i.e. disposal of waste
products via saliva secretion into the mouth.
Especially mucin-rich saliva from the sublingual
and submandibular glands preferentially binds toxins [@doddsHealthBenefits2005],
and since these glands are located closest to the lower incisors, they may be the
most effective target for these studies.
This has yet to be systematically
tested in archaeological dental calculus.
Because we homogenised samples from multiple teeth of an indiviual, we were unable
to test the effect of oral biogeography.
It is also possible that resident microflora within
biofilms contribute to alkaloid breakdown and that the presence of caffeine and
nicotine metabolites following direct ingestion can be explained by this pathway.
However, the literature on
biofilm biodegradation of alkaloids is limited, and *in vitro* studies have only
found minimal contributions by certain oral bacteria in isolation
[@sunMetabolomicsEvaluation2016; @cogoVitroEvaluation2008]; it is possible that
a larger role is played by oral bacteria within larger, more metabolically active
communities, e.g. biofilms [@takahashiOralMicrobiome2015].

<!-- How this engineering will facilitate future research -->

@@

<!-- limitations of the platform show which questions would be unanswered -->
Since we targeted individuals with moderate-to-large calculus deposits,
it is likely a biased sample. The presence of calculus may increase risk of premature death
[@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
be associated with dental calculus build-up) is a risk-factor for
respiratory diseases, if periodontal and respiratory pathogens
enter the blood stream
[@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
In our sample, the percentage of CMS
(`r scales::percent(mean(explore_select$CMS, na.rm = T), accuracy = 0.1)`)
is lower than in another (more representative) male sample (44.1%)
[@casnaUrbanizationRespiratory2021], and the caries percentage is similarly lower
in our sample (`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`) than
a more representative sample (22.9%) [@lemmersMiddenbeemster2013].  
We used the presence/absence of a pipe notch and concurrent detection of tobacco
as a crude estimate of the accuracy of the method, which we found to be around
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
This is a very rough estimate, as the presence of a pipe notch is likely not a
perfect indicator of whether or not someone consumed tobacco.
Dental calculus is also
more transient than for example bone, as it can be mechanically removed,
intentionally or unintentionally, during life<!--cite-->. This would remove all
trace of the alkaloids consumed prior to its removal, and provide a blank slate
for newly formed calculus.  
Quantitation of the detected compounds may have limited value in archaeological
samples, and will greatly affect correlations related to concentration.
The detected quantity of a compound will depend on the quantity in dental
calculus during life, which is largely dependent on quantity of consumption,
how often the
calculus was disrupted/removed, and inter-individual factors such as calculus
formation, and metabolic breakdown of the compound
[@mackiePreservationMetaproteome2017; @velskoAncientDental2022a].
Following burial, compound stability over
time will play a large role, as will microbial degradation of compounds by bacteria
and funghi in soil [@liuNicotinedegradingMicroorganisms2015], as well as
the soil environment, such as temperature, pH, and oxygen availability
[@lindholstLongTerm2010].  
Perhaps the main limitation, which is also true for many other analyses on
archaeological materials, is that detecting the absence of a compound is not possible.
The absence of a compound is
not evidence of its absence, which complicates the interpretation of our results.
We have attempted to minimise errors occurring due to this limitation by including
a relatively large sample of individuals. Although given the relatively low
detection rate seen in tobacco, this remains a major limitation, and will likely
been compounded by increasing antiquity of the samples.
The replication of samples increases the evidence of absence if the compound
was absent in both batches.

Future studies should explore how sampling from various types of teeth
and their position in the mouth effects the probability of a compound becoming
entrapped in dental calculus.
This may also be related to properties within the oral cavity, as well as chemical
properties of the compounds, which facilitate or reduce the incorporation-potential,
and which incorporation pathways are more likely for a given compound.  
We only targeted drugs that were included in the forensic toxicological
screenings, and therefore only covered a limited number of the potential compounds
that could be of interest for exploring past diets and medicinal treatments. The
list of targeted compounds can be expanded as we discover more potential targets
based on which specific compounds/metabolites are more likely to be incorporated
and preserved in dental calculus.  
We found a strong correlation between nicotine and cotinine, but
a lesser correlation between caffeine and theophylline. This is likely because
caffeine and theophylline have multiple plausible sources in our population, but
more research is needed on the relationship between parent compounds and their
metabolites in dental calculus.  
There is an increasing interest in using oral fluid as a means of detecting alkaloids
in living
individuals due to the non-invasive nature of the testing compared to blood
and urine sampling [@coneSalivaTesting1993; @valenDetermination212017].
These *in vivo* studies are a valuable
source of method validation and can help determine the feasibility of detecting
certain alkaloids in oral fluid and, subsequently, dental calculus. Archaeologists,
though, will likely be responsible for exploring dental calculus specific
incorporation and retention of alkaloids, as well as their long-term preservation
in the burial environment. Here, *in vitro* studies
may provide valuable insights<!--shameless self-citation?-->.

<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
While a major limitation is the uncertainty surrounding whether or not a compound
is actually absent, the power of the method is in detecting food and other items
that are consumed in a manner that does not allow macroscopic detection. Taking
tobacco consumption as an example; while pipe
notches are a useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, with others including chewing, drinking, cigars,
and snuff [@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
was mainly practised by males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018],
so chromatographic methods, such as the one presented here,
are suitable for exploring tobacco consumption in an entire population, rather
than a trivial subset of past populations.<!--admittedly, this study did just that...-->
